## Chapter 2: Regression and model validation

### Analysis

XXX TODO:

* Describe the dataset, mention where it can be found. https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt and http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt
* Describe its structure and dimensions
* Show a graphical overview of the data and show summaries of the variables
  in the data.
* Describe and interpret the outputs, commenting on the distributions of the
  variables and the relationships between them.
* Choose three variables as explanatory variables and fit a regression
  model where exam points is the target (dependent) variable.  Show a
  summary of the fitted model and comment and interpret the results.
  Explain and interpret the statistical test related to the model
  parameters.  If an explanatory variable in your model does not have
  a statistically significant relationship with the target variable,
  remove the variable from the model and fit the model again without
  it.
* Using a summary of your fitted model, explain the relationship between
  the chosen explanatory variables and the target variable (interpret the
  model parameters).  Explain and interpret the multiple R squared of the model.
* Produce the following diagnostic plots: Residuals vs. Fitted values,
  Normal QQ-plot and Residuals vs. Leverage.  Explain the assumptions of
  the model and interpret the validity of those assumptions based on the
  diagnostic plots.

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly.
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

### Code and output for the data wrangling exercise

Here is code and output from data/create_learning2014.R:

```{r code=readLines("data/create_learning2014.R")}
```

### Summary of the work done this week and lessons learned

* I went through the DataCamp exercises.  The platform was simple and
  easy to use, even though this would not have been my preferred mode
  of learning.  I would rather have read a description of the R
  philosophy and how main commands work and then approached it as a
  programming language.  Now rather than learning the basics we are
  forced to learn snippets that may be useful in themselves, but we
  don't learn to understand what options and commands really mean or
  what the philosophy and basic concepts are.
* I am becoming increasingly frustrated with the MOOC platform.  This
  is the most confusing online course I have participated in so far.
  All the others were on Moodle, so I don't know if the problem is how
  the material is organized or in the platform.  I'm finding it hard
  to navigate and hard to figure out what exactly we are supposed to
  do.  The exercises themselves are not hard.
* Thus far I have not yet seen anything that would cause me to switch
  from python to R for plotting or regressions.


```{r}
date()
```

Here we go again...
