## Chapter 6: Analysis of longitudinal data

Let's start by loading some libraries we might use.
```{r}
library(ggplot2)
library(GGally)
library(corrplot)
library(dplyr)
library(tidyr)
# Set default figure size
knitr::opts_chunk$set(fig.width=12, fig.height=8)
```

### Data wrangling

The data wrangling for Exercise 6 is performed by the
``data/meet_and_repeat.R`` script in my github repository, as
specified in the instructions.  It can be found
[here](https://github.com/tatuylonen/IODS-project/blob/master/data/meet_and_repeat.R).

### Analysis

First, let's load the data sets created by the data wrangling script.
```{r}
RATS <- read.csv("data/meet_RATS.csv", row.name=1)
RATSL <- read.csv("data/meet_RATSL.csv", row.name=1)
BPRS <- read.csv("data/meet_BPRS.csv", row.name=1)
BPRSL <- read.csv("data/meet_BPRSL.csv", row.name=1)

# Convert ID, Group, treatment, subject fields to Factor
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

str(RATS)
str(RATSL)
str(BPRS)
str(BPRSL)
```

#### (1) Implement the analyses of Chapter 8 of MABS using the RATS data

Ok, so it looks like basically we need to perform the analyses we did
on Datacamp, but with the datasets swapped.

```{r}
# First, look at the data
names(RATSL)
str(RATSL)
summary(RATSL)
glimpse(RATSL)

# Display the data graphically
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

We can see that rats in Group 1 had consistently lower weight than
those in Groups 2 and 3.  However, there were only eight rats in Group
1, four in Group 2, and four in Group 3.  These numbers are so low
that making any statistical analyses based on them are fraught with
uncertainty.  That said, we are tasked with analyzing this data.

We can also see that in each group the rats are gaining weight as time passes.

Let's see how the picture changes if we standardize the data for each
data point (keeping in mind that the number of observations is too low
to do this with any accuracy).

```{r}
# Standardize Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight)) / sd(Weight)) %>%
  ungroup()
str(RATSL)

# Plot again with standardized Weight
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")
```

We can now see clearly that some rats are growing faster than the
others.  Where the line goes down, it means that the rat is gaining
wait slower than the mean of the group, or might even be losing
weight.

Let's now plot the group means and their error bars.  We should note
that it is impossible to calculate variance with any accuracy from
just four observations, and correspondingly the magnitude of the error
bars is only a very rough estimate that shouldn't be given much
statistical significance.

```{r}
# Add mean and standard error computations.  Note that the n in standard error
# computation should be the number of samples used to compute that mean,
# not the total number of samples in the whole dataset for all Group/Time
# combinations (the Datacamp exercise is in error).
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight) / sqrt(length(Weight))) %>%
  ungroup()

# Note: In view of the warning about regrouping above, verify that groups
# are correct inside summarize
dim(RATSL)
X <- RATSL %>% group_by(Group, Time) %>% summarize(cnt=length(Weight)) %>% ungroup()
X$cnt
# Yes, this matches what we see in the earlier plots - 8, 4, 4 rats in each
# group, for each timepoint.  So looks like summarize() worked correctly
# despite the weird warning.

# Let's glimpse at the data with timepoint means and standard errors.
# We leave aside the question whether standard error is reliable with
# just 4 or 8 observations in each grouping...  (it ISN'T very
# reliable with so few observations)
glimpse(RATSS)

# Plot the mean profiles for each Group, with error bars
ggplot(RATSS,
  aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")
```

Let's then check for utliers using a boxplot.  (We should note, however, that
the low number of observations makes any outlier detection very unreliable.)

```{r}
# Quick check for outliers using a boxplot
RATSB <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSB)
ggplot(RATSB, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```

We can see that there is a possible outlier in Group 2.  However,
given that there are only four rats in this group (which is already
too little data), any variance estimate (and thus standard error and
the quartiles) is highly suspect, and thus we cannot really justify
removing any individuals from the data.  They might not be outliers
at all if we had a bit more data, and being already data-deprived,
we would be even more so with only three rats.

Let's then see if there is a statistically significant difference
between the groups.  We'll use the dataset we created with
the mean for each group, but for the reasons above we did not do outlier
elimination.  There is a compication here, as the t test only works between
exactly two groups.  For the purposes of this analysis, we'll lumk groups
2 and 3 together and compare them againt Group 1.  Let's add an auxiliary
field for this.

```{r}
# Create a helper variable indicating whether the observation is in Group 1
RATSB$isGroup1 <- RATSB$Group == "1"

# Perform a t-test between being in group 1 and not being in group 1
t.test(mean ~ isGroup1, data=RATSB, var.equal=TRUE)

# Add a baseline from Time 1
RATSB$baseline <- RATS$WD1

# Fit a linear model with the mean as the response
fit <- lm(mean ~ baseline + isGroup1, data=RATSB)
fit

# Compute an analysis of variance table for the fitted model with anova()
anova(fit)
```

We can see that the data suggests the baseline (WD1) to be a very
significant predictor of the group mean weight and being in Group 1
being significant at 95% level.  However, even ignoring the
unsuitability of the hypothesis testing method testing method for our
case (the observations are not independent), we had so few data points
that the means are inaccurate, and this significance prediction was
based on the inaccurately computed means.  This adds to uncertainty
beyond what was considered by hypothesis testing, and we probably
should not consider being in Group 1 as significant given the baseline
weight.

#### (2) Implement the analyses of Chapter 9 of MABS uising the BPRS data

XXX to be continued
